---
title: "TidyTuesday_23_11_7"
author: "LBE"
date: "2023-11-06"
output: md_document
---

# Week 45

## setup
```{r}
library(tidyverse)
library(tidytuesdayR)
library(ggpubr)
library(ggimage)
library(png)
library(ggtext)
library(patchwork)

tuesdata <- tidytuesdayR::tt_load(2023, week = 45)
house <- tuesdata$house
```

## Explore

Let's explore today's data and see what we have! Only instructions this week are to stay clear of **causation**, fair enough!

```{r}
house

(plot1 <- ggplot(data = house, aes(x=year, y=candidatevotes, fill = party)) +
  geom_point(shape = 21))

```

Ok, looks like independents really clutter up these data - let's remove them to get a better picture

```{r}

(plot1 <- ggplot(data = subset(house, party=="DEMOCRAT"|party=="REPUBLICAN"), aes(x=year, y=candidatevotes, fill = party)) +
  geom_point(shape = 21))
```

Hang on, need to do some reading about US house elections... OK, so to win the house, you need 216 seats. Each seat belongs to a district and majority votes in each district decides seat.

Perhaps it would be interesting to see whether third party candidates have the power to tip the house? so compare the years when independents or writeins made up the difference for each party controlling th house?

```{r}
#overall data
house_data <- house %>% 
  group_by(year, state, district) %>% 
 mutate(ranked = rank(desc(candidatevotes)),
        won = if_else(ranked == 1, 1, 0)) %>%
  group_by(year, party) %>% 
  summarise(seats = sum(won)) 

# only R vs D
house_data_twoparty <- house %>% 
  filter(party == "DEMOCRAT" |party == "REPUBLICAN") %>% 
    group_by(year, state, district) %>% 
  select(year, state, district, party, candidatevotes)

#only other
house_data_other <- house %>% 
  filter(party != "DEMOCRAT" & party != "REPUBLICAN") %>% 
  select(year, state, district, party, candidatevotes) %>% 
  dplyr::group_by(year, state, district) %>% 
  summarise(candidatevotes = sum(candidatevotes)) %>% 
  mutate (party = "OTHER") 

#seats awarded to each if recieved all 'other' votes
house_data_spoiler <- as.data.frame(house_data_other) %>% 
  add_row(house_data_twoparty) %>%
  pivot_wider(names_from = party, values_from = candidatevotes, values_fn = sum, values_fill = 0) %>% 
  mutate(D = (DEMOCRAT + OTHER),
         R = (REPUBLICAN + OTHER),
         Dwin = if_else(D>R, 1, 0),
         Rwin = if_else(R>D, 1, 0)) %>% 
  group_by(year) %>% 
  summarise(REPUBLICAN = sum(Rwin),
            DEMOCRAT = sum(Dwin)) %>% 
  pivot_longer(2:3, names_to = 'party', values_to = "seats")

#which years are a spoiler year, and in what direction?  
house_data_comp <-  house_data %>% 
  filter(party == "DEMOCRAT" |party == "REPUBLICAN") %>% 
  right_join(house_data_spoiler, by = c("year", "party")) %>% 
  mutate(spoil = if_else(seats.y>seats.x, seats.y,NA),
         imp = if_else(spoil==1 & seats.x<=216 & seats.y>216, 1, NA)) 
```


##let's plot!
```{r}
ggplot(house_data, aes(x = year, y = seats)) +
  geom_point(shape = 21, size =2, alpha = 0.8, fill = "grey")+
  geom_point(data = subset(house_data, party == "REPUBLICAN"|party== "DEMOCRAT"), aes(x=year, y=seats, fill = party), shape = 21, size =2)+
    geom_hline(yintercept = 216)+
  geom_point(data = house_data_comp, aes(x=year, y=spoil, colour = party), shape =8)+
    theme(legend.position="none")+
  scale_fill_manual(values = c("blue", "red"))+
  scale_colour_manual(values = c("blue", "red"))+

  theme_classic()
      
```

Ok, so now we have the data we want, let's make it pretty! I feel like we might not need to include the others, as it kind of just clutters up the image?

``` {r}

# Import the image
library(magick)
img <- image_read("../data/watercolourpaper.jpg") %>% 
  image_blur(3, 3) %>%
  image_colorize(opacity = 70, color = "beige")
image_write(img, "../data/bg.jpg")

caption <- ("<br>*Spoiling* is when smaller parties split the vote from the two dominant parties.<br> <br>Here we look at the *unlikely* scenario that each party recieved **ALL** of the votes cast for parties other than the <span style = 'color:blue;'>**Democrats**</span> or <span style = 'color:red;'>**Republicans**</span>, and whether that would have an influence on who controls the House of Representatives. <br> <br> There has been **19 elections** since 1976 where spoiler votes would have changed the number of seats awarded to each party (imaginary seats marked with **'*'**), but in **no case was this enough to change which party control the house**<br>")

(spoiler_elections <- ggplot(house_data, aes(x = year, y = seats)) +
  geom_line(data = subset(house_data, party == "REPUBLICAN"|party== "DEMOCRAT"), aes(x=year, y=seats, colour = party), alpha = 0.3, size = 1 )+
   geom_point(data = subset(house_data, party == "REPUBLICAN"|party== "DEMOCRAT"), aes(x=year, y=seats, fill = party), shape = 21, linewidth =1, alpha = 0.7)+
    geom_hline(yintercept = 216, alpha = 0.5)+
  geom_point(data = house_data_comp, aes(x=year, y=spoil, colour = party), shape =8)+
    annotate("text", x=1979, y=225, label = "House Majority")+
    # inset_element(p=img2, top = 1, left = 1, bottom = 1, right = 1)+
   scale_fill_manual(values = c("blue", "red"))+
  scale_colour_manual(values = c("blue", "red"))+
    labs(x = "Election Year", y = "House of Representative Seats",
         title = "Is spoiling an issue in the House of Representatives?",
         subtitle = caption)+
    coord_cartesian(ylim = c(140, 300), 
                      clip = 'off') +
   theme_classic()+
   theme(legend.position = "none",
         panel.border = element_rect(linewidth = 0.5, fill=NA),
         plot.margin = unit(c(1,1,1,1), "cm"),
         plot.title = element_text(face = "bold"),
         plot.subtitle = element_textbox_simple())
     
 )
spoiler_elections
TT_week45 <-  ggbackground(spoiler_elections, "../data/bg.jpg")
```
## Final Figure
There we go, final figure!  

**What I like:** Answers the questions, in a simple way  

**What I don't like:** It's a bit boring?  

``` {r Week45_figure}
ggsave("../figures/week_45.jpg", plot = TT_week45)

TT_week45

```
